Overview and Summary of the First Workshop on Sustainable Software for Science: Practice and Experiences (WSSSPE1) \note{\scriptsize Authors, please add/check middle initials and affiliations}

### 2014-04-11 WSSSPE1_report/WSSSPE1_summary_report.tex
edits in communities text

Christopherson reports on the application of Steps 1-3 to a
computational modeling framework developed in the 1990s. Step 1 was realized as
a 2-day, in person specifications meeting and code walkthrough. Step 2 was a
5-day hackathon to develop working code, and Step 3 was a 3-day hackathon to refine
the code based on new requirements. The team worked on small, low-risk units of
code. It was challenging, revealed unanticipated obstacles, programmers had to
work together, and experimentation was encouraged.

The paper summarized recommendations to others wishing to engage in this or a
similar process, starting small and gradually building toward more complex
objectives. This is consistent with Agile development. Refactor before adding
new functionality. Approach development as a learning experience. Welcome
experimentation, and treat mistakes as a natural part of the learning process.
Repeat Step 1 activities before all hackathons to develop consensus before
coding. This allows coding to be the focus of subsequent hackathons. In higher
risk situations, provide additional time for Step 1 activities. The
Christopherson team recommends a minimum of two months. Ensure any newcomers
receive some form of orientation prior to the hackathon, such as a code
walkthrough or system documentation. Co-locate rather than collaborating
remotely whenever feasible.

Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can provide a user-friendly
entry to complex cyberinfrastructure. The paper opens with a description of the
explosion of use of just a few gateways. Over 3.5 years, more than 7,000
biologists have run phylogenetic codes on supercomputers using the\toolname{CIPRES}
Science Gateway. In 1 year, over 120 scientists from 50 institutions used the
UltraScan Science Gateway, increasing the sophistication of analytical
ultracentrifugation experiments worldwide. The new Neuroscience Gateway (\toolname{NSG})
registered 100+ users who used 250,000 CPU hours in only a few months.

Gateways, however, need to keep operational costs low and can often make use of
common components. Science Gateway Platform as a Service (\toolname{SciGaP}) delivers
middleware as a hosted, scalable third-party service while domain developers
focus on user interfaces and domain-specific data in the gateway. The
middleware handles things like authentication, application installation and
reliable execution and help desk support.

One key to \toolname{SciGaP} is its openness. While based on the \toolname{Apache Airavata} project
and the \toolname{CIPRES} Workbench Framework, community contributions are encouraged
because of its open source, open governance and open operations policies. The
goal is robust, sustainable infrastructure with a cycle of development that
improves reliability and prioritizes stakeholder requirements. The project is
leveraging Internet2's \toolname{Net+} mechanisms for converting \toolname{SciGaP} and its gateways
into commodity services.

Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and challenges with the large
nanoHUB.org community. The authors define community as a ``body of persons of
common and especially professional interests scattered through a larger
society.'' This makes support challenging because of the diversity of
viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging
particular types of use.

nanoHUB's 10-year history with over 260,000 users annually provides a lot of
data to analyze. 4000 resources contributed by 1000 authors. nanoHUB serves
both the research and education community and the contribution model allows
researchers to get their codes out into the community and in use in education
very rapidly. The primary software challenges are twofold - support for the
HUBzero framework and challenges related to the software contributed by the
community.

The group has learned that community contributions are maximized with a
tolerant licensing approach. HUBzero uses an LGPLv3 license so contributors can
create unique components and license as they choose. If they make changes to
source code, the original license must be maintained for redistribution. As far
as contributed resources, these must be open access, but not necessarily open
source. This allows contributors to meet the requirements of their institutions
and funding agencies. Quality is maintained via user ratings. Documentation is
encouraged and nanoHUB supplies regression test capabilities, but the user
community provides ratings, poses questions and contributes to wishlists and
citation counts - all of which incentivize code authors.

Terrel~\cite{Terrel_WSSSPE} describes support for the Python scientific community through
two major efforts - the SciPy conference and the NumFOCUS foundation. Reliance
on software in science has driven a huge demand for development, but this
development is typically done as a side effort and often in a rush to publish
without documentation and testing. While created by academics, software support
often falls to industrial institutions. SciPy brings together industry,
government, and academics to share their code and experience in an open
environment.

NumFOCUS promotes open, usable scientific software while sustaining the
community. Specific activities include educational programs, collaborative
research tools and documentation and promotion of high-level languages,
reproducible scientific research, and open-code development. Governance of the
non-profit is a loose grantor-grantee relationship with projects allowing for
monies to be placed in the groups accounts. This has raised money to hire
developers for open code development, maintain testing machines, organize the
PyData conference series, and sponsor community members to attend conferences.

Software sustainability relies on contributions from all sectors of the user
community. Together SciPy and NumFOCUS support these sectors. By maximizing
contributions growing the user base they help develop and mature Python.

L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project which was started in 1996 by
participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh
and thorns, core and module, model - a community-oriented framework that allows
researchers to easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain dormant,
becoming active only when parameters and simulation data dictate. Users can
use modules written by others or can write their own modules without changing
other code. The community has grown and diversified beyond the original science
areas.

The paper points out 4 keys to sustaining the community - modular design,
growing a collaborative community, career paths and credit. In modular design,
the Cactus project went far beyond standard practices of APIs. Domain specific
languages (DSLs) allow decoupling of components - for example I/O, mesh
refinement, PAPI counters, and boundary conditions abstracted from science
code. In academia, publications are the main currency of credit. Because the
project connects code developments to science, the work is publishable and
modules are citable. Because of the open source, modular approach, programmers
can see the impact of their contributions and often continue work after
graduation. Career paths remain a challenge, however. Tasks that are essential
from a software engineering perspective are often not rewarded in academia. The
best programmers in a science environment often have multidisciplinary
expertise. That also is not rewarded in academia.

Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF software institute effort to build a
community of those creating science gateways or web portals for science. These
gateways, as described in some of the other papers in this section, can be
quite capable, having a remarkable impact on some parts of science.

This paper mentioned challenges highlighted by other papers in this section -
mainly the conflict between funding for research vs infrastructure and the
challenges around getting academic credit for infrastructure. Because the
authors have studied many projects, they've also observed how development is
often done in an isolated hobbyist environment. Developers are unable to take
advantage of similar work done by others, isolation even when projects have
common needs. But often projects struggle for sustainable funding because they
provide infrastructure to conduct research and many times only the research is
funded. Gateways also may start as small group research project, taking off in
popularity once they go live without any long term plans for sustainability or
without resources in the project to plan for such. Subsequent disruptions in
service can limit effectiveness and test the limits of the research community's
trust. The impact of gateways can be increased substantially if we understand
what makes them successful.

Recommendations from an early study of successful gateways include the
following. Leadership and management teams should design governance to
represent multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the technical and
domain-related issues, consider sustainability and measure success early and
often. Successful projects recognize the benefits and costs of hiring a team of
professionals, demonstrate credibility through stability and clarity of
purpose, leverage the work of others and plan for flexibility. Successful
projects identify an existing community and understand the communities' needs
before beginning. Projects adapt as the communities' needs evolve. For funders,
the lifecycle of technology projects must be considered. Solicitations should
be designed to reward effective planning, recognize the benefits and
limitations of both technology innovation and reuse, expect adjustments during
the production process, copy effective models from other industries and
sectors, and encourage partnerships that support gateway sustainability.

Through a business incubator type approach, the institute plans to provide a
variety of services that could be shared amongst projects. Consultation and
resources on topics such as business plan development, software engineering
practices, software licensing options, usability, security and project
management as well as a software repository and hosting service will be
available. Experts will be available for multi-month assignments to help
research teams build their own gateways, teaching them how to maintain and add
to the work after the collaboration ends. Forums, symposia and an annual
conference will connect members of the development community. A modular,
layered framework that supports community contributions and allows developers
to choose components will be delivered and finally workforce development
activities will help train the next generation for careers in this cross-
disciplinary area and build pools of institutional expertise that many projects
can leverage. Shared services and forums can dramatically reduce the cost of
building and sustaining gateways. Workforce development can encourage
technology professionals to remain in the sciences.

@@

\subsubsection{Measuring community}
How might a project know when it has built a sustainable community?  How might an outsider be able to assess the activity and sustainability of that community?  These questions have been partially addressed in the literature.  For example, a key metric in online communities in general is the cross-over point where there are more external contributors than internal ones. Richard Millington's book ``Buzzing Communities'' does an excellent job of outlining these measures, drawing on online communities research in an accessible manner~\cite{millington_buzzing_2012}. Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}.

>>>>

Christopherson reports on the application of Steps 1-3 to a
computational modeling framework developed in the 1990s. Step 1 was
a 2-day, in-person specifications meeting and code walkthrough. Step 2 was a
5-day hackathon to develop working code, and Step 3 was a 3-day hackathon to refine
the code based on new requirements. The team worked on small, low-risk units of
code. It was challenging, revealed unanticipated obstacles, programmers had to
work together, and experimentation was encouraged.
%
The paper recommended: start small and gradually building toward more complex
objectives, consistent with Agile development; develop consensus before
coding, by repeating step 1 before all hackathons; ensure newcomers
receive orientation prior to the hackathon, such as a code
walkthrough or system documentation; and co-locate collaborators
whenever feasible.

%Refactor before adding
%new functionality. Approach development as a learning experience. Welcome
%experimentation, and treat mistakes as a natural part of the learning process.
%Repeat Step 1 activities before all hackathons to develop consensus before
%coding. This allows coding to be the focus of subsequent hackathons. In higher
%risk situations, provide additional time for Step 1 activities. The
%Christopherson team recommends a minimum of two months. 
%Ensure any newcomers
%receive some form of orientation prior to the hackathon, such as a code
%walkthrough or system documentation. Co-locate rather than collaborating
%remotely whenever feasible.

Pierce et al.~\cite{Pierce2_WSSSPE} described how science gateways can provide a user-friendly
entry to complex cyberinfrastructure. For example, more than 7,000
biologists have run phylogenetic codes on supercomputers using the \toolname{CIPRES}
Science Gateway in 3 1/2 years. Over 120 scientists from 50 institutions used the
\toolname{UltraScan Science Gateway} in one year, increasing the sophistication of analytical
ultracentrifugation experiments worldwide. The new Neuroscience Gateway (\toolname{NSG})
registered 100+ users who used 250,000 CPU hours in only a few months.

Gateways, however, need to keep operational costs low and can often make use of
common components, such as authentication, application installation and
reliable execution and help desk support. Science Gateway Platform as a Service (\toolname{SciGaP}) delivers
middleware as a hosted, scalable third-party service while domain developers
focus on user interfaces and domain-specific data in the gateway. 
While \toolname{SciGaP} is based on the \toolname{Apache Airavata} project
and the \toolname{CIPRES} Workbench Framework, community contributions are encouraged
by its open source, open governance and open operations policies. The
goal is robust, sustainable infrastructure with a cycle of development that
improves reliability and prioritizes stakeholder requirements. The project is
leveraging Internet2's \toolname{Net+} mechanisms for converting \toolname{SciGaP} and its gateways
into commodity services.

Zentner et al.~\cite{Zentner_WSSSPE} describes experiences and challenges with the large
\toolname{nanoHUB.org} community, where community is defined as a ``body of persons of
common and especially professional interests scattered through a larger
society.'' Support is challenging because of the diversity of
viewpoints and needs. The group constantly examines its policies to determine
whether they are indirectly alienating part of the community or encouraging
particular types of use.
%
\toolname{nanoHUB}'s 10-year history with over 260,000 users annually provides a lot of
data to analyze. 4000 resources contributed by 1000 authors. \toolname{nanoHUB} serves
both the research and education community and the contribution model allows
researchers to get their codes out into the community and in use in education
very rapidly. The primary software challenges are twofold - support for the
\toolname{HUBzero} framework and challenges related to the software contributed by the
community.

The group has learned that community contributions are maximized with a
tolerant licensing approach. \toolname{HUBzero} uses an LGPLv3 license so contributors can
create unique components and license as they choose. If they make changes to
source code, the original license must be maintained for redistribution. As far
as contributed resources, these must be open access, but not necessarily open
source. This allows contributors to meet the requirements of their institutions
and funding agencies. Quality is maintained via user ratings. Documentation is
encouraged and nanoHUB supplies regression test capabilities, but the user
community provides ratings, poses questions and contributes to wishlists and
citation counts, all of which incentivize code authors.

Terrel~\cite{Terrel_WSSSPE} describes support for the Python scientific community through
two major efforts: the SciPy conference and the NumFOCUS foundation. 
Since software sustainability relies on contributions from all sectors of the user
community, these efforts support these sectors, and help develop and mature Python.
%
The reliance
on software in science has driven a huge demand for development, but this
development is typically done as a side effort and often in a rush to publish
without documentation and testing. While the software is often created by academics, software support
can fall to industrial institutions. SciPy brings together industry,
government, and academics to share their code and experience in an open
environment.
%
NumFOCUS in a non-profit that promotes open, usable scientific software while sustaining the
community through educational programs; collaborative
research tools and documentation; and promotion of high-level languages,
reproducible scientific research, and open-code development. Governance
is a loose grantor-grantee relationship with projects, allowing funds
to be placed in the groups' accounts. This has raised money to hire
developers for open code development, maintain testing machines, organize the
PyData conference series, and sponsor community members to attend conferences.

L\"{o}ffler et al.~\cite{Loffler_WSSSPE} describes the Cactus project,
 which was started in 1996 by
participants in USA Binary Black Hole Alliance Challenge. Cactus has a flesh (core)
and thorns (modules) model, a community-oriented framework that allows
researchers to easily work together with reusable and extendable software
elements. Modules are compiled into an executable and can remain dormant,
becoming active only when parameters and simulation data dictate. Users can
use modules written by others or can write their own modules without changing
other code. The community has grown and diversified beyond the original science
areas.

The paper points out 4 keys to sustaining the community: modular design,
growing a collaborative community, career paths and credit. In modular design,
the Cactus project went far beyond standard practices of APIs. Domain specific
languages (DSLs) allow decoupling of components - for example I/O, mesh
refinement, PAPI counters, and boundary conditions abstracted from science
code. In academia, publications are the main currency of credit. Because the
project connects code developments to science, the work is publishable and
modules are citable. Because of the open source, modular approach, programmers
can see the impact of their contributions and often continue work after
graduation. Career paths remain a challenge, however. Tasks that are essential
from a software engineering perspective are often not rewarded in academia. The
best programmers in a science environment often have multidisciplinary
expertise. This also is not rewarded in academia.

Wilkins-Diehr et al.~\cite{Wilkins-Diehr_WSSSPE} describes an NSF software institute effort to build a
community of those creating science gateways for science. These
gateways, as described in some of the other papers in this section, can be
quite capable and can have strong scientific impact.
Challenges are similar to those highlighted by other papers in this section:
the conflict between funding for research vs infrastructure and the
challenges around getting academic credit for infrastructure. 
%
The authors observe that  development is
often done in an isolated hobbyist environment. Developers are unable to take
advantage of similar work done by others, isolation even when projects have
common needs. But often projects struggle for sustainable funding because they
provide infrastructure to conduct research and many times only the research is
funded. Gateways also may start as a small group research project, taking off in
popularity once they go live, without any long term plans for sustainability.
Subsequent disruptions in
service can limit effectiveness and test the limits of the research community's
trust. %The impact of gateways can be increased substantially if we understand
%what makes them successful.

Recommendations from an early study of successful gateways include:
1) Leadership and management teams should design governance to
represent multiple strengths and perspectives, plan for change and turnover in
the future, recruit a development team that understands both the technical and
domain-related issues, consider sustainability and measure success early and
often. 2) Projects should hire a team of
professionals, demonstrate credibility through stability and clarity of
purpose, leverage the work of others, and plan for flexibility. 3) Projects should
identify an existing community and understand the communities' needs
before beginning, then adapt as the communities' needs evolve. 4) Funders,
should consider the technology project lifecycle, and design solicitations 
to reward effective planning, recognize the benefits and
limitations of both technology innovation and reuse, expect adjustments during
the production process, copy effective models from other industries and
sectors, and encourage partnerships that support gateway sustainability.

%Through a business incubator type approach, the institute plans to provide a
%variety of services that could be shared amongst projects. Consultation and
%resources on topics such as business plan development, software engineering
%practices, software licensing options, usability, security and project
%management as well as a software repository and hosting service will be
%available. Experts will be available for multi-month assignments to help
%research teams build their own gateways, teaching them how to maintain and add
%to the work after the collaboration ends. Forums, symposia and an annual
%conference will connect members of the development community. A modular,
%layered framework that supports community contributions and allows developers
%to choose components will be delivered and finally workforce development
%activities will help train the next generation for careers in this cross-
%disciplinary area and build pools of institutional expertise that many projects
%can leverage. Shared services and forums can dramatically reduce the cost of
%building and sustaining gateways. Workforce development can encourage
%technology professionals to remain in the sciences.

@@

\subsubsection{Measuring community}
How might a project know when it has built a sustainable community?  How might an outsider be able to assess the activity and sustainability of that community?  These questions have been partially addressed in the literature.  For example, a key metric in online communities in general is the cross-over point where there are more external contributors than internal ones. Richard Millington's book ``Buzzing Communities'' does an excellent job of outlining these measures, drawing on online communities research in an accessible manner~\cite{millington_buzzing_2012}. Some participants in the workshop have since prepared online materials outlining current and future practices in measurement of scientific software and its ecosystem\footnote{\url{https://sites.google.com/site/scientificsoftwaremetrics/}}. \katznote{please change to cite}
